---
title: "Area of Applicability: geovalidate vs CAST"
subtitle: "Verifying the Python re-implementation against Meyer's R reference"
author: "geovalidate"
format:
  html:
    code-fold: false
    toc: true
    embed-resources: true
execute:
  warning: false
  message: false
---

# Setup

This document compares `geovalidate.area_of_applicability` (Python) with
`CAST::aoa` (R), the original reference implementation from
Meyer & Pebesma (2021). Both calculate the Area of Applicability (AOA)
for a model trained on `X_train` and applied to `X_test`.

```{r}
#| label: setup-r
library(CAST)
library(reticulate)

# Point reticulate at the conda env that ships geovalidate. The
# RETICULATE_PYTHON env var (set when invoking quarto) takes precedence;
# we fall back to an explicit miniforge path so the document is still
# reproducible when rendered without env vars.
py_bin <- Sys.getenv("RETICULATE_PYTHON",
                     unset = "~/miniforge/envs/geovalidate/bin/python")
use_python(py_bin, required = TRUE)
```

```{python}
#| label: setup-py
import numpy
from geovalidate import area_of_applicability
print("geovalidate area_of_applicability imported")
```

# Synthetic dataset

We generate a 5-feature training cluster centred on the origin, and a
test set composed of (a) in-domain points drawn from the same
distribution, and (b) out-of-domain points shifted by 8 standard
deviations. AOA should flag the in-domain points as applicable and the
shifted ones as not.

```{r}
#| label: data-r
set.seed(0)
n_train <- 200
n_in    <- 30
n_out   <- 20
n_feat  <- 5

X_train <- matrix(rnorm(n_train * n_feat), n_train, n_feat)
y_train <- X_train[, 1] + 2 * X_train[, 2] + 0.1 * rnorm(n_train)

X_in    <- matrix(rnorm(n_in  * n_feat, mean = 0), n_in,  n_feat)
X_out   <- matrix(rnorm(n_out * n_feat, mean = 8), n_out, n_feat)
X_test  <- rbind(X_in, X_out)
is_in_domain <- c(rep(TRUE, n_in), rep(FALSE, n_out))

colnames(X_train) <- paste0("x", seq_len(n_feat))
colnames(X_test)  <- paste0("x", seq_len(n_feat))

cat("train:", nrow(X_train), "x", ncol(X_train),
    "  test:", nrow(X_test), "x", ncol(X_test), "\n")
```

# CAST::aoa (R reference)

We pass a uniform feature weight vector so the comparison is purely
about the distance + normalisation + threshold pipeline (model-based
permutation weights are non-deterministic, which would muddy the side-
by-side).

```{r}
#| label: cast-uniform
w_uniform <- rep(1.0 / n_feat, n_feat)
names(w_uniform) <- colnames(X_train)

cast_res <- aoa(
  newdata     = as.data.frame(X_test),
  train       = as.data.frame(X_train),
  weight      = as.data.frame(rbind(w_uniform)),
  variables   = colnames(X_train),
  CVtest      = NULL,
  CVtrain     = NULL,
  verbose     = FALSE
)

cast_di       <- as.numeric(cast_res$DI)
cast_aoa_mask <- as.logical(cast_res$AOA)
cast_thresh   <- attr(cast_res, "aoa_stats")$threshold
if (is.null(cast_thresh)) cast_thresh <- cast_res$parameters$threshold

cat("CAST threshold:", round(cast_thresh, 4), "\n")
cat("CAST DI range :", round(range(cast_di), 3), "\n")
cat("CAST applicable: in-domain =",
    round(mean(cast_aoa_mask[is_in_domain]) * 100, 1), "%, ",
    "OOD =", round(mean(cast_aoa_mask[!is_in_domain]) * 100, 1), "%\n")
```

# geovalidate.area_of_applicability (Python)

Same data, same uniform weights, called via reticulate.

```{python}
#| label: geo-uniform
X_train = numpy.asarray(r.X_train)
X_test  = numpy.asarray(r.X_test)
is_in   = numpy.asarray(r.is_in_domain, dtype=bool)

geo = area_of_applicability(
    X_test, X_train,
    feature_weights="uniform",
    threshold="tukey",
    return_diagnostics=True,
)

geo_di     = geo.dissimilarity_index
geo_mask   = geo.applicable
geo_thresh = geo.cutpoint

print(f"geovalidate threshold: {geo_thresh:.4f}")
print(f"geovalidate DI range : [{geo_di.min():.3f}, {geo_di.max():.3f}]")
print(f"geovalidate applicable: in-domain = {geo_mask[is_in].mean()*100:.1f}%, "
      f"OOD = {geo_mask[~is_in].mean()*100:.1f}%")
```

# Side-by-side comparison

## Cut-off

```{r}
#| label: compare-cutoff
geo_thresh <- py$geo_thresh
cat(sprintf("CAST threshold       : %.4f\n", cast_thresh))
cat(sprintf("geovalidate threshold: %.4f\n", geo_thresh))
cat(sprintf("relative difference  : %.2f%%\n",
            100 * abs(cast_thresh - geo_thresh) / cast_thresh))
```

## Dissimilarity index, per test point

```{r}
#| label: compare-di-table
geo_di <- as.numeric(py$geo_di)
ord <- order(cast_di)
data.frame(
  i               = seq_along(cast_di)[ord],
  in_domain       = is_in_domain[ord],
  cast_DI         = round(cast_di[ord], 3),
  geovalidate_DI  = round(geo_di[ord], 3),
  ratio_geo_cast  = round(geo_di[ord] / cast_di[ord], 3)
) |> head(15)
```

```{r}
#| label: compare-di-plot
#| fig-width: 5.5
#| fig-height: 5.5
plot(cast_di, geo_di,
     xlab = "CAST DI", ylab = "geovalidate DI",
     pch  = ifelse(is_in_domain, 20, 4),
     col  = ifelse(is_in_domain, "steelblue", "tomato"),
     main = "Per-point DI: CAST vs geovalidate")
abline(0, 1, lty = 2, col = "grey50")
legend("topleft",
       legend = c("in-domain", "out-of-domain", "y = x"),
       pch    = c(20, 4, NA),
       lty    = c(NA, NA, 2),
       col    = c("steelblue", "tomato", "grey50"),
       bty    = "n")
```

## AOA agreement

```{r}
#| label: compare-mask
geo_mask  <- as.logical(py$geo_mask)
cast_mask <- cast_aoa_mask

agree <- geo_mask == cast_mask
cat(sprintf("AOA mask agreement: %d / %d (%.1f%%)\n",
            sum(agree), length(agree), 100 * mean(agree)))

cat("\nConfusion (rows = CAST, cols = geovalidate):\n")
print(table(CAST = cast_mask, geovalidate = geo_mask))
```

# Why DIs may differ in magnitude

CAST and geovalidate share the same recipe -- weighted feature
distances, normalisation by mean within-training distance, threshold
from the Tukey rule on the training-DI distribution -- but the
implementations diverge on small details:

* **Normalisation constant.** CAST divides by the mean of the
  full pairwise training-distance matrix (including the zero diagonal,
  via its internal mean). The Python implementation here uses the
  upper triangle only. This shifts every DI by a constant scale
  factor.
* **Weight application.** CAST multiplies the feature columns by
  `weight` (a single-row data.frame keyed by variable name);
  geovalidate multiplies the columns by a 1-D numpy array. Numerically
  identical when the column orders match.
* **Threshold history.** Original nanophyto/abil had two
  off-by-100-percentile bugs in the Tukey + percentile threshold paths
  that geovalidate fixes (see `_aoa.py` for the FIXED-relative-to
  comments). CAST is the reference here, so disagreement on the
  threshold scale should be attributed to that fix, not to CAST.

What we should expect to match: the **AOA boolean mask** (modulo any
points sitting almost exactly on the threshold), and the **rank
ordering** of test points by DI (they're computing the same distances
and dividing by constants, so the orderings are identical even when
the absolute DI values differ).
